---
title: "autodiff"
author: "Mathieu"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    toc: true
    toc_depth: 3

---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
library(stringr)
knitr::opts_chunk$set(echo = TRUE)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(fig.path='Figs/')
```

# Benchmark description

## Macro benchmark

- flu dataset from treetime_validation
- 5000 iterations
- model:
  - JC69
  - time tree
  - constant population size
  - exp(1000) prior on strict clock
  - OneOnX prior on population size

## Micro benchmark

- flu dataset from treetime_validation
- 1000 replicates
- treelikelihood:
  - input: branch lengths
  - output: log likleihood
  - derivatives wrt branch lengths
- coalescent:
  - input: node heights and pop size
  - output: log likelihood
  - derivatives wrt constant pop size and node heights (not ratios)
- ratio_transform:
  - input: ratios/root height
  - output: internal node heights
  - derivatives wrt ratios
- ratio_transform_jacobian:
  - input: internal node heights
  - output: log determinant of the Jacobian
  - derivatives wrt internal node heights

Programs:

- torchtree: pure pytorch
- bitorch: torchtree with BITO (analytic derivatives of tree likelihood calculated with BEAGLE)
- torchtree32: torchtree with single precision
- phylojax: pure JAX
- treeflow: pure treeflow
- physher: pure C (analytic derivatives)
- phylostan: Stan language

\newpage
# Results
## Macro benchmark

### Runtime against dataset size
```{r include=FALSE}
df.macro.raw = read.csv("resultsd/macro/macro.csv", sep = ',')
max_iter = 5000 # number of iterations in the benchmark
```

```{r}
df.macro = df.macro.raw %>%
  filter(iters == max_iter | program == 'treeflow') %>%
  group_by(program, size) %>% summarize(time = min(time)) %>% ungroup()

ggplot(df.macro, aes(x = size,
                     y = time,
                     color = program)) +
  geom_line() +
  geom_point() +
  xlab("Number of sequences") +
  ylab("Time in seconds")
```

\newpage
### Runtime against dataset size without phylojax
```{r}
df.macro = df.macro.raw %>%
  filter(iters == max_iter | program == 'treeflow') %>%
  filter(program != 'phylojax') %>%
  group_by(program, size) %>% summarize(time = min(time)) %>% ungroup()

ggplot(df.macro, aes(x = size,
                     y = time,
                     color = program)) +
  geom_line() +
  geom_point() +
  xlab("Number of sequences") +
  ylab("Time in seconds")
```

### Runtime against dataset size up to 500 sequences
```{r}
df.macro = df.macro.raw %>%
  filter(iters == max_iter | program == 'treeflow') %>%
  filter(size <= 500) %>%
  filter(program != 'phylojax') %>%
  group_by(program, size) %>% summarize(time = min(time)) %>% ungroup()

ggplot(df.macro, aes(x = size,
                     y = time,
                     color = program)) +
  geom_line() +
  geom_point() +
  xlab("Number of sequences") +
  ylab("Time in seconds")
```


\newpage
```{r include=FALSE}
df.micro = read.csv("resultsd/micro/micro.csv", sep = ',')

df.micro[df.micro$program == 'treeflow' &
           df.micro$function. == 'constant_coalescent', ]$function. = 'coalescent'

df.micro$function.[df.micro$function. == 'treelikelihoodJC69'] = 'treelikelihood'

# Use the minimum time out of the x replicates
df.micro = df.micro %>% group_by(program, size, precision, function., mode, JIT) %>%
  summarize(time = min(time)) %>% ungroup()

df.micro[df.micro$precision == 32,]$program = 'torchtree32'

df.micro = df.micro %>% filter(function. != 'ratio_transform_jacobian2' &
                                 function. != 'ratio_transform2')

df.micro = df.micro %>%
  mutate(program = ifelse(
    function. == 'treelikelihoodAnalytic',
    'jax-analytic',
    program
  ))  %>%
  mutate(function. = ifelse(
    function. == 'treelikelihoodAnalytic',
    'treelikelihood',
    function.
  ))
df.micro = df.micro %>% filter(program != 'jax-analytic')

# add first and other evaluation times
df.micro.jax = df.micro %>% filter(program == 'phylojax')
df.micro.jax.1 = df.micro.jax %>% filter(grepl('1', mode)) %>%
  mutate(mode = str_sub(mode, 1, nchar(mode) - 1))
df.micro.jax.rest = df.micro.jax %>% filter(!grepl('1', mode))
df.micro.jax.all = df.micro.jax.1 %>%
  full_join(df.micro.jax.rest,
            by = c("program", "size", "precision", "function.", "JIT", "mode")) %>%
  mutate(time.x = time.x + time.y) %>%
  rename(time = time.x) %>% select(!time.y)

df.micro = rbind(
  df.micro %>% filter(program != 'phylojax'),
  df.micro.jax.all,
  df.micro.jax.1 %>% mutate(program = 'phylojax1')
)
```

\newpage

## Micro benchmark
### Log likelihood evalutation
```{r}
micro.like = df.micro %>% filter(mode == 'evaluation')

ggplot(micro.like, aes(
  x = size,
  y = time,
  color = program,
  linetype = JIT
)) +
  geom_line() +
  geom_point() +
  facet_wrap( ~ function., scales = "free", nrow = 2) +
  xlab("Number of sequences") +
  ylab("Time in seconds")
```

\newpage

### Gradient evaluation
```{r}
micro.grad = df.micro %>% filter(mode == 'gradient')


ggplot(micro.grad, aes(
  x = size,
  y = time,
  color = program,
  linetype = JIT
)) +
  geom_line() +
  geom_point() +
  facet_wrap( ~ function., scales = "free", nrow = 2) +
  xlab("Number of sequences") +
  ylab("Time in seconds")
```

\newpage

### Log likelihood evaluation wihtout phylojax
```{r}
micro.grad = df.micro %>% filter(mode == 'evaluation') %>%
  filter(!grepl('phylojax', program))

ggplot(micro.grad, aes(
  x = size,
  y = time,
  color = program,
  linetype = JIT
)) +
  geom_line() +
  geom_point() +
  facet_wrap( ~ function., scales = "free", nrow = 2) +
  xlab("Number of sequences") +
  ylab("Time in seconds")
```

\newpage

### Gradient evaluation wihtout phylojax
```{r}
micro.grad = df.micro %>% filter(mode == 'gradient') %>%
  filter(!grepl('phylojax', program))

ggplot(micro.grad, aes(
  x = size,
  y = time,
  color = program,
  linetype = JIT
)) +
  geom_line() +
  geom_point() +
  facet_wrap( ~ function., scales = "free", nrow = 2) +
  xlab("Number of sequences") +
  ylab("Time in seconds")
```

\newpage

### Gradient evaluation log scale
```{r}
dd = df.micro %>% filter(mode == 'gradient') %>% filter(program != 'phylojax')
ggplot(dd, aes(
  x = size,
  y = time,
  color = program,
  linetype = JIT
)) +
  geom_point() +
  stat_smooth(formula = y ~ x,
              method = "lm",
              se = FALSE) +
  scale_x_continuous(trans = 'log2') +
  scale_y_continuous(trans = 'log2') +
  facet_wrap( ~ function., scales = "free", nrow = 2)
```

\newpage

```{r}
fit_log_log <- function(df, m) {
  fits = df %>% filter(mode == m) %>%
    mutate(size = log(size), time = log(time)) %>%
    group_by(program, JIT, function., mode) %>%
    do(model = lm(time ~ size, data = .)) %>% ungroup()
  
  fits$intercept = sapply(fits$model, function(x)
    x$coefficients[[1]])
  fits$slope = sapply(fits$model, function(x)
    x$coefficients[[2]])
  fits = fits %>% select(!model)
  fits
}
```

\newpage

### Log likelihood evaluation log scale
```{r}
dd.log = fit_log_log(df.micro, 'evaluation')

kable(dd.log %>% arrange(function., slope))
```

\newpage

### Gradient evaluation log scale
```{r}
dd.log.grad = fit_log_log(df.micro, 'gradient')

kable(dd.log.grad %>% arrange(function., slope))
```

\newpage

### Only bitorch and physher
```{r}
dd = df.micro %>%  filter(program %in% c('physher', 'bitorch')) %>%
  filter(function. != 'coalescent')

ggplot(dd, aes(
  x = size,
  y = time,
  color = program,
  linetype = mode
)) +
  geom_line() +
  geom_point() +
  facet_wrap( ~ function., scales = "free") +
  xlab("Number of sequences") +
  ylab("Time in seconds")
```

\newpage

### Ratio of gradient and likelihood execution time
```{r}
micro.join = micro.like %>%
  full_join(micro.grad,
            by = c("program", "size", "precision", "function.", "JIT")) %>%
  mutate(ratio = time.y / time.x) %>% filter(!grepl('phylojax', program))

ggplot(micro.join, aes(
  x = size,
  y = ratio,
  color = program,
  linetype = JIT
)) +
  geom_line() +
  geom_point() +
  facet_wrap( ~ function., scales = "free") +
  xlab("Number of sequences") +
  ylab("Ratio gradient/evaluation execution times")
```